# Analyse Contamination {#contamination-analysis}

```{r setup, include=FALSE}

```

The heartpiece of the study is [this sql script](https://github.com/grst/BioQC_GEO_analysis/blob/master/db/views/contamination.sql) which identifies heterogenous samples. In this section we will go through the different analysis steps performed in the script. 

```{r, echo=FALSE}
knitr::include_graphics("db/design/contamination-views.svg")
```

## Input Tables
* `BIOQC_SELECTED_SAMPLES`: the samples resulting from the [selection process](#selecting samples) including metadata
* `BIOQC_TISSUE_SET`: the [manual mapping](#tissues-signatures) of tissues to signatures ("expected signatures")
* `BIOQC_RES`: the p-values for each sample and signature (pre-filtered for having a p-value < 0.05)

## Processing Steps explained
### BIOQC_SELECTED_SAMPLES_TSET
We join `BIOQC_SELECTED_SAMPLES` on `BIOQC_TISSUE_SET` to attach the tissue group and the respective signatures
to each sample.

Example: The table rows for the colon sample `GSM1234` could look like: 
```
GSM       TISSUE    TISSUE_SET  TGROUP    EXP_SIGNATURE   EXP_SIGNATURE_NAME 
GSM1234   colon     gtex_all    colon     543             Colon
GSM1234   colon     gtex_solid  intestine 543             Colon
GSM1234   colon     gtex_solid  intestine 544             Jejunum
```



### BIOQC_RES_TSET
We join the BioQC results (pvalue for each signature and sample) from `BIOQC_RES` with `BIOQC_TISSUE_SET` to map the signatures back to their tissue group. 

Example: For a result
```
GSM       SIGNATURE     PVALUE
GSM1234   544           1e-10
GSM1234   543           1e-5
```
the table rows could look like
```
GSM       FOUND_SIG     PVALUE    FOUND_SIG_NAME  FOUND_TGROUP  TISSUE_SET
GSM1234   544           1e-10     Jejunum         jejunum       gtex_all
GSM1234   544           1e-10     Jejunum         intestine     gtex_solid
GSM1234   543           1e-5      Colon           colon         gtex_all
GSM1234   543           1e-5      Colon           intestine     gtex_solid
```

### BIOQC_TISSUE_ENRICHMENT
We now match the found onto the expected tissue groups by joining `BIOQC_SELECTED_SAMPES_TSET` with `BIOQC_RES_TSET` on `GSM` and `TISSUE_SET`. 
We calculate the *Enrichment Ratio* as $\log_{10}(p_{expected\_sig} / p_{found\_sig})$. 
This table containes *only* the contamined samples; samples where no 'unexpected' signature is found are excluded.

For the above example tables, we would get: 
```
GSM       TISSUE    TISSUE_SET  TGROUP    EXP_SIGNATURE   EXP_SIGNATURE_NAME  FOUND_SIG  FOUND_SIG_NAME  ENRICHMENT_RATIO
GSM1234   colon     gtex_all    colon     543             Colon               543        Colon           0.0
GSM1234   colon     gtex_all    colon     543             Colon               544        Jejunum         5.0
...
```

We filter out rows where the found signature is in the list of *expected sigantures* for the respective tissue group:
```
GSM       TISSUE    TISSUE_SET  TGROUP    EXP_SIGNATURE   EXP_SIGNATURE_NAME  FOUND_SIG  FOUND_SIG_NAME  ENRICHMENT_RATIO
GSM1234   colon     gtex_all    colon     543             Colon               544        Jejunum         5.0
...
```

### BIOQC_TISSUE_ENRICHMENT2
combine expected signature by taking the minimal enrichment ratio
for each expected signature. 

Example:
```
GSM    TISSUE    TGROUP    EXPECTED  FOUND         ENRICHMENT_RATIO
GSM888 jejunum   intestine colon     liver_fetal   12
GSM888 jejunum   intestine colon     liver         8
GSM888 jejunum   intestine jejunum   liver_fetal   5
GSM888 jejunum   intestine jejunum   liver         4
```

will be combined into
```
GSM              TGROUP    EXPECTED  FOUND         ENRICHMENT_RATIO  RANK
GSM888           intestine           liver_fetal   5                 1
GSM888           intestine           liver         4                 1
```

if multiple infiltrating tissues are found, a rank is calculated.


### BIOQC_CONTAM_STATS
Add year, country and organism to each sample from `BIOQC_TISSUE_ENRICHMENT2`. 


### BIOQC_TISSUE_MIGRATION
Add *origin* and *destination* tissue groups for each sample by mapping the found signatures back to tissue groups. 

Example:
```
GSM       TISSUE_SET  ORIGIN  FOUND_SIG  FOUND_SIG_NAME  ENRICHMENT_RATIO   DESTINATION
GSM1234   gtex_all    colon   544        Jejunum         5                  jejunum
```


